19Q1

### 2019-05-15 19Q1.Rmd
update 19Q1 Rmd

Annotation files for mutation status, copy number, and gene expression of the CCLE and CTRP datasets (Consortium and Consortium 2015, Barretina et al. 2012) were downloaded from the [DepMap Data Portal](https://depmap.org/portal/download/). Note that the group that conducted the GDSC screen also collected their own genomic annotation information for mutation status, copy number, and gene expression, but we did not use it.

@@

@@

  ifelse(g2p$Feature.Names == "KRAS G12G" | g2p$Feature.Names == "KRAS G13G", "synonymous",

@@

DepMap 19Q2 fusion calls:
```{r fusion_calls, eval=FALSE}
fusions <- read.delim("./data_munging/19Q2/CCLE_fusions.csv.gz", header = TRUE, sep = ",")
fusions$LeftGene <- gsub(" .*", "", fusions$LeftGene)
fusions$RightGene <- gsub(" .*", "", fusions$RightGene)
fusions$ <- gsub(" .*", "", fusions$RightGene)
```

@@

19Q1, CCLE RNAseq gene expression data (log2(TPM+1)):
```{r gene_expression, eval=FALSE}
ge <- read.delim("./../htscreens_giant_files/CCLE_depMap_19Q1_TPM_20190305.csv.gz", header = TRUE, sep = ",", check.names = FALSE)
# Remove Entrez gene IDs from colnames, format is "Gene (Entrez ID)"
colnames(ge) <- gsub(" .*", "", colnames(ge))
colnames(ge)[1] <- "Broad_ID"
# Melt
ge_melt <- melt(ge, id.vars = "Broad_ID", measure.vars = colnames(ge)[2:ncol(ge)], variable.name = "Gene", value.name = "TPM")
ge_melt$TPM <- log2(ge_melt$TPM + 1)
ge_melt <- unfactorize(ge_melt)

@@

```{r omics_functions}
MergeOmicsG2P <- function(database, outfile) {
  # Merge mutation status, CN, and GE omics data with drugscreen results and metadata
  if(database %in% c("CCLE", "CTRP", "GDSC")) {
    filt <- auc %>% filter(Database == database) %>% select(Cell_line, Histology, TCGA_classification, CCLE_Name, Broad_ID, Drug, AUC_IC50)
    print(paste0("Processing ", database, " data."))
    colnames(filt)[colnames(filt) == "AUC_IC50"] <- "Score"

    # Get all G2P gene-cell line combinations
    grid <- expand.grid("Cell_line" = unique(filt$Cell_line), "Gene" = g2p_genes)
    data <-  merge(filt, grid, by = "Cell_line", all.y = TRUE)
    print("Merged grid.")
    # Merge metadata
    data <-  merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = c("Cell_line", "Histology", "CCLE_Name", "Broad_ID"), all.y = TRUE)
    ## Add columns for G2P drug-gene and drug-gene-lineage sorting
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$G2P_Drug_Gene <- ifelse(data$Drug_Gene %in% g2p_druggenes, "Yes", "No")
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    data$G2P_Drug_Gene_Lin <- ifelse(data$Drug_Gene_Lin %in% g2p_dgls, "Yes", "No")
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df_ptmuts, by = c("Gene", "Broad_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }
  if(database == "CRISPR") {
    print(paste0("Processing ", database, " data."))
    # Melt CRISPR dataset for merging
    data <- melt(crispr_all, id.vars = "Broad_ID", measure.vars = colnames(crispr_all)[2:ncol(crispr_all)], variable.name = "Gene", value.name = "Score")
    # Get all G2P gene-cell line combinations
    data <-  filter(data, Gene %in% g2p_genes)
    # Merge cell line metadata
    data <- merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = "Broad_ID", all.y = TRUE)
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "Broad_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }

@@

  # Merge GE data
  data <- merge(data, ge_filt, by = c("Gene", "Broad_ID"), all.x = TRUE)
  print("Merged gene expression data.")

  # Merge CN data
  data <- merge(data, cn_filt, by = c("Gene", "Broad_ID"), all.x = TRUE)
  print("Merged copy number data.")

@@

```{r merge_filter_drug_data, eval=FALSE}
# Load in Pozdeyev et al. (2016) data
ccle_all <- read.delim("./../htscreens_giant_files/pozdeyev_data/CCLE_drug_response_metrics.csv.gz", sep = ",")
ctrp_all <- read.delim("./../htscreens_giant_files/pozdeyev_data/CTRP_drug_response_metrics.csv.gz", sep = ",")
gdsc_all <- read.delim("./../htscreens_giant_files/pozdeyev_data/GDSC_drug_response_metrics.csv.gz", sep = ",")

@@

# Omics metadata
maf_df_ptmuts <- unique(filter(maf_df, Gene %in% g2p_genes & PointMutation == 1) %>% select(-c(Variant_Classification, Reference_Allele, Tumor_Seq_Allele1, PointMutation)))
ge_filt <- filter(ge_melt, Gene %in% g2p_genes)
cn_filt <- filter(cn_melt, Gene %in% g2p_genes)

ccle_ptmuts <- MergeOmicsG2P(database = "CCLE", outfile = "./data_munging/rds/ccle_data_19Q1_g2pgenes_ptmuts.rds")
ctrp_ptmuts <- MergeOmicsG2P(database = "CTRP", outfile = "./../htscreens_giant_files/ctrp_data_19Q1_g2pgenes_ptmuts.rds")
gdsc_ptmuts <- MergeOmicsG2P(database = "GDSC", outfile = "./data_munging/rds/gdsc_data_19Q1_g2pgenes_ptmuts.rds")
```

@@

# Testing
test_ccle <- ccle_ptmuts %>% count(Gene, Drug)
test_ctrp <- ctrp_ptmuts %>% count(Gene, Mutation_Status)
test_gdsc <- gdsc_ptmuts %>% count(Gene, Mutation_Status)
```

>>>>

Annotation files for mutation status, copy number, and gene expression of the CCLE, GDSC, and CTRP datasets (Consortium and Consortium 2015, Barretina et al. 2012) were downloaded from the [DepMap Data Portal](https://depmap.org/portal/download/). Note that the group that conducted the GDSC screen also collected their own genomic annotation information for mutation status, copy number, and gene expression, but we did not use it.

@@

# Manually assign biomarker_types based on a combination of biomarker_type derived from the JSON string and patterns in the Feature.Names

@@

  ifelse(g2p$Feature.Names == "KRAS G12G" | g2p$Feature.Names == "KRAS G13G", "Silent",

@@

DepMap 19Q2 fusion calls derived using the [STAR-Fusion pipeline](https://github.com/STAR-Fusion/STAR-Fusion/wiki). Column descriptions:

- *JunctionReads*: the number of RNA-Seq fragments containing a read that aligns as a split read at the site of the putative fusion junction
- *SpanningFrags*: the number of RNA-Seq fragments that encompass the fusion junction such that one read of the pair aligns to a different gene than the other paired-end read of that fragment
    - Those predictions that have very few JunctionReads and/or SpanningReads are going to be enriched for false positives. Note, depending on the site of the fusion breakpoint and length of the reads, it may not be possible to have SpanningFragments and all evidence may show up in the form of JunctionReads
- *FFPM (fusion fragments per million total reads)*: normalized measure of the fusion-supporting RNA-seq fragmentS; a filter of 0.1 sum FFPM (meaning at least 1 fusion-supporting RNA-seq fragment per 10M total reads) tends to be effective at excluding fusion artifacts, and is the current default for filtering fusions from the final output
- *LargeAnchorSupport*: whether there are split reads that provide 'long' (set to length of 25 bases) alignments on both sides of the putative breakpoint
    - Those fusions supported only by split reads (no spanning fragments) and lack LargeAnchorSupport are often highly suspicious and tend to be false positives
    - Those with LargeAnchorSupport are labeled as 'YES_LDAS' (where LDAS = long double anchor support)
- *SpliceType*: whether the proposed breakpoint occurs at reference exon junctions as provided by the reference transcript structure annotations (ex. gencode)
- *LeftBreakEntropy* and *RightBreakEntropy*: the Shannon entropy of the 15 exonic bases flanking the breakpoint.
    - The maximum entropy is 2, representing highest complexity, and the lowest would be zero (involving a 15 base mononucleotide run)
    - Low entropy sites should generally be treated as less confident breakpoints.

```{r fusion_calls_19Q2, eval=FALSE}
fusions <- read.delim("./data_munging/19Q2/CCLE_fusions.csv.gz", header = TRUE, sep = ",")
fusions$LeftGene <- gsub(" .*", "", fusions$LeftGene)
fusions$RightGene <- gsub(" .*", "", fusions$RightGene)
fusions$X.FusionName <- gsub("--", "-", fusions$X.FusionName)
fusions <- select(fusions, DepMap_ID, X.FusionName)
fusions <- as.data.frame(lapply(fusions, function(x) gsub("IGH@|IGH-@", "IGH", x)))
colnames(fusions) <- c("Broad_ID", "FusionPair")
```

@@

19Q1, CCLE RNAseq gene expression data:
```{r gene_expression, eval=FALSE}
ge <- read.delim("./../htscreens_giant_files/CCLE_depMap_19Q1_TPM_20190305.csv.gz", header = TRUE, sep = ",", check.names = FALSE)
# Remove Entrez gene IDs from colnames, format is "Gene (Entrez ID)"
colnames(ge) <- gsub(" .*", "", colnames(ge))
colnames(ge)[1] <- "Broad_ID"
# Melt
ge_melt <- melt(ge, id.vars = "Broad_ID", measure.vars = colnames(ge)[2:ncol(ge)], variable.name = "Gene", value.name = "TPM")
ge_melt$TPM <- log2(ge_melt$TPM + 1)
ge_melt <- unfactorize(ge_melt)

@@

```{r omics_functions}
MergeOmicsG2P <- function(database, maf_df, ge_df, cn_df, outfile) {
  # Merge mutation status, CN, and GE omics data with drugscreen results and metadata
  if(database %in% c("CCLE", "CTRP", "GDSC")) {
    filt <- auc %>% filter(Database == database) %>% select(Cell_line, Histology, TCGA_classification, CCLE_Name, Broad_ID, Drug, AUC_IC50)
    print(paste0("Processing ", database, " data."))
    colnames(filt)[colnames(filt) == "AUC_IC50"] <- "Score"

    # Get all G2P gene-cell line combinations
    grid <- expand.grid("Cell_line" = unique(filt$Cell_line), "Gene" = g2p_genes)
    data <-  merge(filt, grid, by = "Cell_line", all.y = TRUE)
    print("Merged grid.")
    # Merge metadata
    data <-  merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = c("Cell_line", "Histology", "CCLE_Name", "Broad_ID"), all.y = TRUE)
    ## Add columns for G2P drug-gene and drug-gene-lineage sorting
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$G2P_Drug_Gene <- ifelse(data$Drug_Gene %in% g2p_druggenes, "Yes", "No")
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    data$G2P_Drug_Gene_Lin <- ifelse(data$Drug_Gene_Lin %in% g2p_dgls, "Yes", "No")
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "Broad_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }
  if(database == "CRISPR") {
    print(paste0("Processing ", database, " data."))
    # Melt CRISPR dataset for merging
    data <- melt(crispr_all, id.vars = "Broad_ID", measure.vars = colnames(crispr_all)[2:ncol(crispr_all)], variable.name = "Gene", value.name = "Score")
    # Get all G2P gene-cell line combinations
    data <-  filter(data, Gene %in% g2p_genes)
    # Merge cell line metadata
    data <- merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = "Broad_ID", all.y = TRUE)
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "Broad_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }

@@

  # Merge GE data
  data <- merge(data, ge_df, by = c("Gene", "Broad_ID"), all.x = TRUE)
  print("Merged gene expression data.")

  # Merge CN data
  data <- merge(data, cn_df, by = c("Gene", "Broad_ID"), all.x = TRUE)
  print("Merged copy number data.")

@@

**MERGE WITHOUT G2P AND PTMUT FILTER (5/15)**
```{r merge_filter_drug_data, eval=FALSE}
# Load in Pozdeyev et al. (2016) data
ccle_all <- read.delim("./../htscreens_giant_files/pozdeyev_data/CCLE_drug_response_metrics.csv.gz", sep = ",")
ctrp_all <- read.delim("./../htscreens_giant_files/pozdeyev_data/CTRP_drug_response_metrics.csv.gz", sep = ",")
gdsc_all <- read.delim("./../htscreens_giant_files/pozdeyev_data/GDSC_drug_response_metrics.csv.gz", sep = ",")

@@

ccle <- MergeOmicsG2P(database = "CCLE", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./data_munging/rds/ccle_data_19Q1.rds")
# ctrp <- MergeOmicsG2P(database = "CTRP", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./../htscreens_giant_files/ctrp_data_19Q1.rds")
# gdsc <- MergeOmicsG2P(database = "GDSC", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./data_munging/rds/gdsc_data_19Q1.rds")

# Omics metadata
# maf_df_ptmuts <- unique(filter(maf_df, Gene %in% g2p_genes & PointMutation == 1)
# ge_filt <- filter(ge_melt, Gene %in% g2p_genes)
# cn_filt <- filter(cn_melt, Gene %in% g2p_genes)
# ccle_ptmuts <- MergeOmicsG2P(database = "CCLE", maf_df = maf_df_ptmuts, ge_df = ge_filt, cn_df = cn_filt, outfile = "./data_munging/rds/ccle_data_19Q1_g2pgenes_ptmuts.rds")
# ctrp_ptmuts <- MergeOmicsG2P(database = "CTRP", maf_df = maf_df_ptmuts, ge_df = ge_filt, cn_df = cn_filt, outfile = "./../htscreens_giant_files/ctrp_data_19Q1_g2pgenes_ptmuts.rds")
# gdsc_ptmuts <- MergeOmicsG2P(database = "GDSC", maf_df = maf_df_ptmuts, ge_df = ge_filt, cn_df = cn_filt, outfile = "./data_munging/rds/gdsc_data_19Q1_g2pgenes_ptmuts.rds")
```

```{r}
ccle_missense <- filter(ccle, Variant_Classification == "Missense_Mutation")
ccle_missense_grid <- expand.grid("Gene" = unique(ccle_missense$Gene), "Broad_ID" = unique(ccle_missense$Broad_ID))
ccle_missense_grid$Variant_Classification <- NA
ccle_missense <- merge(ccle_missense, ccle_missense_grid, by = c("Gene", "Broad_ID", "Variant_Classification"), all = TRUE)
```

@@

# Testing
test_ccle <- ccle_ptmuts %>% count(Drug)
test_ctrp <- ctrp_ptmuts %>% count(Drug)
sum(!is.na(str_match(test_ctrp$Drug, "ratio")))
test_gdsc <- gdsc_ptmuts %>% count(Drug)
```
